---
title: "1.2 CoPPRA PTM QC plots"
output:
  html_document:
    toc: true
    number_sections: true
date: "`r format(Sys.Date())`"
---

Run after 1.1. This file reads saved QC results and creates PNG/PDF plots in `QC/plots`; it does not read evidence or repeat quantification.

```{r setup}
source("CoPPRA_config_SEA.R", local = TRUE)
library(ggplot2)
library(pheatmap)
library(grid)
```


# Read saved QC results



```{r read_qc}
paths_SEA <- stage_paths_SEA("QC")
plot_folder_SEA <- paths_SEA$plots
qc_SEA <- readRDS(file.path(paths_SEA$data, "QC_plot_data_SEA.rds"))
list2env(setNames(qc_SEA, paste0(names(qc_SEA), "_SEA")), envir = environment())
theme_set(theme_bw(base_size = 14) + theme(
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  axis.title = element_text(size = 16), axis.text = element_text(size = 14),
  strip.text = element_text(size = 16, face = "bold")))
```

# Peptide intensity distributions



```{r intensity_plot, fig.width=18, fig.height=7}
int_plot_SEA <- merge(
  pep_run_SEA[, .(
    Raw.file,
    Log2_intensity = log2(Peptide_intensity)
  )],
  runs_SEA[, .(Raw.file, Condition, Bio_rep)],
  by = "Raw.file"
)

p_intensity_SEA <- ggplot(
  int_plot_SEA,
  aes(
    x = factor(Bio_rep, levels = 1:3),
    y = Log2_intensity,
    fill = Condition
  )
) +
  geom_boxplot(width = 0.72, outlier.size = 0.2) +
  facet_wrap(~Condition, nrow = 1) +
  scale_fill_manual(values = condition_colors_SEA, drop = FALSE) +
  labs(
    title = "Peptide Intensity Distributions",
    x = "Biological replicate",
    y = "log2 peptide intensity"
  ) +
  theme(legend.position = "none")

print(p_intensity_SEA)
save_plot_SEA(p_intensity_SEA, "QC_peptide_intensity_boxplot", 10, 4)
```

# Coverage



```{r coverage_plot, fig.width=18, fig.height=9}
cov_plot_SEA <- rbindlist(list(
  run_cov_SEA[, .(
    Condition,
    Bio_rep,
    Metric = "Peptides",
    Coverage = Unique_peptides
  )],
  run_cov_SEA[, .(
    Condition,
    Bio_rep,
    Metric = "Protein groups",
    Coverage = Leading_protein_groups
  )]
))
cov_plot_SEA[, Metric := factor(
  Metric,
  levels = c("Peptides", "Protein groups")
)]

p_coverage_SEA <- ggplot(
  cov_plot_SEA,
  aes(
    x = factor(Bio_rep, levels = 1:3),
    y = Coverage,
    fill = Condition
  )
) +
  geom_boxplot(width = 0.72) +
  facet_grid(Metric ~ Condition, scales = "free_y") +
  scale_fill_manual(values = condition_colors_SEA, drop = FALSE) +
  labs(
    title = "Peptide and Protein Coverage",
    x = "Biological replicate",
    y = "Distinct identifications"
  ) +
  theme(legend.position = "none")

print(p_coverage_SEA)
save_plot_SEA(p_coverage_SEA, "QC_coverage_boxplot", 10, 5)
```

# Missingness



```{r missingness_plot, fig.width=18, fig.height=7}
p_missingness_SEA <- ggplot(
  run_cov_SEA,
  aes(
    x = factor(Raw.file, levels = raw_order_SEA),
    y = 100 * Peptide_missingness,
    fill = Condition
  )
) +
  geom_col() +
  scale_fill_manual(values = condition_colors_SEA, drop = FALSE) +
  labs(
    title = "Peptide-level Missingness by Technical Run",
    x = "Technical run",
    y = "Missing peptide values (%)",
    fill = "Condition"
  ) +
  theme(
    axis.text.x = element_text(
      size = 12,
      angle = 90,
      hjust = 1,
      vjust = 0.5
    )
  )

print(p_missingness_SEA)
save_plot_SEA(p_missingness_SEA, "QC_peptide_missingness_by_raw_file", 10, 4)
```

# Technical correlations



```{r technical_heatmap, fig.width=20, fig.height=18}
run_annotation_SEA <- data.frame(
  Sample = runs_SEA$Experiment,
  Condition = as.character(runs_SEA$Condition),
  Tech_run = factor(runs_SEA$Tech_rep),
  row.names = runs_SEA$Raw.file
)

make_heatmap_SEA <- function(
    matrix, order, annotation, title, angle, show_names = TRUE) {
  heatmap <- pheatmap(
    matrix[order, order],
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    annotation_row = annotation[order, , drop = FALSE],
    annotation_col = annotation[order, , drop = FALSE],
    annotation_colors = list(Condition = condition_colors_SEA),
    # Annotation names are already shown in the legend; avoid duplicate labels
    # colliding with the 18-point legend and 16-point sample names.
    annotation_names_row = FALSE,
    annotation_names_col = FALSE,
    border_color = NA,
    fontsize = 18,
    fontsize_row = 16,
    fontsize_col = 16,
    show_rownames = show_names,
    show_colnames = show_names,
    angle_col = angle,
    main = title,
    silent = TRUE
  )
  title_row <- which(heatmap$gtable$layout$name == "main")
  heatmap$gtable$grobs[[title_row]]$gp$fontsize <- 20
  heatmap$gtable$layout[title_row, c("l", "r")] <- list(1L, ncol(heatmap$gtable))
  # Repeating the same annotations on both axes can offset pheatmap's legend
  # below the canvas. Align its first visible heading with the top instead.
  legend_row <- which(heatmap$gtable$layout$name == "annotation_legend")
  if (length(legend_row)) {
    legend <- heatmap$gtable$grobs[[legend_row]]
    offset <- grid::unit(1, "npc") - legend$children[[1]]$y
    for (i in seq_along(legend$children)) {
      legend$children[[i]]$y <- legend$children[[i]]$y + offset
    }
    heatmap$gtable$grobs[[legend_row]] <- legend
  }
  heatmap
}

run_heatmap_SEA <- make_heatmap_SEA(
  run_cor_SEA,
  raw_order_SEA,
  run_annotation_SEA,
  "Pearson Correlation among Technical Runs",
  "90",
  FALSE
)

grid.newpage()
grid.draw(run_heatmap_SEA$gtable)
save_plot_SEA(
  run_heatmap_SEA$gtable,
  "QC_Pearson_correlation_technical_runs",
  12,
  11
)
```

# Biological correlations



```{r sample_heatmap, fig.width=12, fig.height=11}
sample_annotation_SEA <- data.frame(
  Condition = as.character(samples_SEA$Condition),
  Biological_replicate = factor(samples_SEA$Bio_rep),
  row.names = samples_SEA$Experiment
)

sample_heatmap_SEA <- make_heatmap_SEA(
  sample_cor_SEA,
  sample_order_SEA,
  sample_annotation_SEA,
  "Pearson Correlation among Samples",
  "45"
)

grid.newpage()
grid.draw(sample_heatmap_SEA$gtable)
save_plot_SEA(
  sample_heatmap_SEA$gtable,
  "QC_Pearson_correlation_samples",
  8,
  7
)
```


# Coverage before and after QC

Protein counts are verified MaxQuant protein groups; peptides are distinct unmodified sequences. Counts include all 45 raw files, with a separate view excluding carriers.

```{r coverage_before_after, fig.width=12, fig.height=7}
coverage_comparison_SEA <- as.data.table(qc_SEA$qc_coverage_before_after)
coverage_comparison_SEA <- coverage_comparison_SEA[Stage %in% c("Before_QC", "After_complete_QC")]
coverage_long_SEA <- melt(coverage_comparison_SEA,
  id.vars = c("Stage", "Scope"),
  measure.vars = c("Evidence_rows", "Peptide_sequences", "Verified_assigned_groups"),
  variable.name = "Metric", value.name = "Count")
coverage_long_SEA[, Stage := factor(Stage,
  levels = c("Before_QC", "After_complete_QC"), labels = c("Before QC", "After QC"))]
coverage_long_SEA[, Metric := factor(Metric,
  levels = c("Evidence_rows", "Peptide_sequences", "Verified_assigned_groups"),
  labels = c("Evidence rows", "Peptide sequences", "Protein groups"))]
coverage_long_SEA[, Scope := factor(Scope,
  levels = c("All_samples", "Non_carrier"), labels = c("All samples", "Non-carrier samples"))]
p_before_after_SEA <- ggplot(coverage_long_SEA, aes(Stage, Count, fill = Stage)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = format(Count, big.mark = ",", scientific = FALSE, trim = TRUE)),
    vjust = -0.35, size = 3.5) +
  facet_wrap(~ Scope + Metric, scales = "free_y", ncol = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.16))) +
  scale_fill_manual(values = c("Before QC" = "#8A99A6", "After QC" = "#287C8E")) +
  labs(title = "Coverage before and after QC", x = NULL, y = "Count") +
  theme(legend.position = "none")
print(p_before_after_SEA)
save_plot_SEA(p_before_after_SEA, "QC_coverage_before_after", 12, 7)
```
