---
title: "4.1 CoPPRA SEA biological concept analysis"
output:
  html_document:
    toc: true
    number_sections: true
date: "`r format(Sys.Date())`"
---

Run after 3.1. Rank proteins using median peptide signed normal scores from limma logFC and raw P values, then run GO BP/MF/CC and KEGG GSEA separately for R, U and RU. Operational CT concepts pass BH-adjusted P < 0.10 in RU and are nonsignificant or unavailable in both single treatments. This is an operational comparison, not an interaction test. KEGG requires internet access. Packages must be installed before running the series.

```{r setup}
source("CoPPRA_config_SEA.R", local = TRUE)
library(dplyr)
library(tidyr)
library(readr)
library(AnnotationDbi)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
```


# Read limma results and set parameters



```{r read_results}
paths_SEA <- stage_paths_SEA("Biological_Concept")
table_folder_SEA <- paths_SEA$tables
minimum_pathway_size_SEA <- 10
maximum_pathway_size_SEA <- 500
adjusted_p_cutoff_SEA <- 0.10
minimum_term_similarity_SEA <- 0.20
show_category_SEA <- 12
set.seed(20260908)
peptide_results_SEA <- copy(readRDS(file.path(output_root_SEA, "Limma", "data",
                                         "Limma_results_SEA.rds"))$peptide_significance)
```

# Calculate protein scores



```{r protein_scores}
signed_z_SEA <- function(log_fc, p_value) {
  # Keep unavailable statistics unavailable; only clamp finite P values at zero.
  p_value <- pmin(pmax(p_value, .Machine$double.xmin), 1)
  sign(log_fc) * stats::qnorm(p_value / 2, lower.tail = FALSE)
}

peptide_results_SEA[, `:=`(
  Ru_score = signed_z_SEA(
    Ru_vs_Control_limma_logFC,
    Ru_vs_Control_pvalue
  ),
  Ul_score = signed_z_SEA(
    Ul_vs_Control_limma_logFC,
    Ul_vs_Control_pvalue
  ),
  RU_score = signed_z_SEA(
    RU_vs_Control_limma_logFC,
    RU_vs_Control_pvalue
  )
)]

protein_score_long_SEA <- peptide_results_SEA[
  !is.na(Proteins) & Proteins != "",
  .(
    ProteinID = trimws(unlist(strsplit(
      Proteins,
      ";",
      fixed = TRUE
    )))
  ),
  by = .(
    Sequence,
    Ru_score,
    Ul_score,
    RU_score
  )
]

protein_score_long_SEA <- unique(protein_score_long_SEA[ProteinID != ""], by = c("Sequence", "ProteinID"))

protein_scores_SEA <- protein_score_long_SEA[, .(
  n_peptides = uniqueN(Sequence),
  R_score = median(Ru_score, na.rm = TRUE),
  U_score = median(Ul_score, na.rm = TRUE),
  RU_score = median(RU_score, na.rm = TRUE)
), by = ProteinID]

protein_scores_SEA[, UniProt := sub("-\\d+$", "", ProteinID)]

cat("Protein accessions with ranking scores:", nrow(protein_scores_SEA), "\n")
```

# Map UniProt to Entrez IDs



```{r protein_mapping}
uniprot_ids_SEA <- unique(protein_scores_SEA$UniProt)

id_map_SEA <- clusterProfiler::bitr(
  uniprot_ids_SEA,
  fromType = "UNIPROT",
  toType = c("ENTREZID", "SYMBOL"),
  OrgDb = org.Hs.eg.db
) %>%
  dplyr::distinct(UNIPROT, ENTREZID, SYMBOL)

protein_scores_mapped_SEA <- protein_scores_SEA %>%
  dplyr::inner_join(
    id_map_SEA,
    by = c("UniProt" = "UNIPROT")
  ) %>%
  dplyr::group_by(ENTREZID) %>%
  dplyr::summarise(
    SYMBOL = paste(sort(unique(SYMBOL)), collapse = ";"),
    UniProt = paste(sort(unique(UniProt)), collapse = ";"),
    n_accessions = dplyr::n_distinct(ProteinID),
    n_peptides = sum(n_peptides),
    R_score = median(R_score, na.rm = TRUE),
    U_score = median(U_score, na.rm = TRUE),
    RU_score = median(RU_score, na.rm = TRUE),
    .groups = "drop"
  )

readr::write_csv(
  id_map_SEA,
  file.path(table_folder_SEA, "uniprot_to_entrez_mapping_SEA.csv")
)

readr::write_csv(
  protein_scores_mapped_SEA,
  file.path(table_folder_SEA, "protein_rank_scores_SEA.csv")
)

mapping_summary_SEA <- tibble::tibble(
  Metric = c(
    "Protein accessions before mapping",
    "Mapped Entrez IDs used for ranking"
  ),
  Value = c(
    nrow(protein_scores_SEA),
    nrow(protein_scores_mapped_SEA)
  )
)

readr::write_csv(
  mapping_summary_SEA,
  file.path(table_folder_SEA, "ranking_mapping_summary_SEA.csv")
)

print(mapping_summary_SEA)
```

# Build ranked lists



```{r ranked_lists}
make_ranked_list_SEA <- function(data, score_column) {
  score <- data[[score_column]]
  names(score) <- as.character(data$ENTREZID)

  score <- score[is.finite(score) & !is.na(names(score))]

  tie_break <- rank(
    names(score),
    ties.method = "first"
  ) / length(score)

  score <- score + tie_break * 1e-12
  sort(score, decreasing = TRUE)
}

ranked_lists_SEA <- list(
  R = make_ranked_list_SEA(protein_scores_mapped_SEA, "R_score"),
  U = make_ranked_list_SEA(protein_scores_mapped_SEA, "U_score"),
  RU = make_ranked_list_SEA(protein_scores_mapped_SEA, "RU_score")
)

rank_summary_SEA <- tibble::tibble(
  Contrast = names(ranked_lists_SEA),
  n_ranked_proteins = vapply(ranked_lists_SEA, length, integer(1)),
  minimum_score = vapply(ranked_lists_SEA, min, numeric(1)),
  maximum_score = vapply(ranked_lists_SEA, max, numeric(1))
)

readr::write_csv(
  rank_summary_SEA,
  file.path(table_folder_SEA, "ranked_list_summary_SEA.csv")
)

print(rank_summary_SEA)
```

# Run GSEA



```{r gsea}
make_readable_gsea_SEA <- function(x) {
  if (is.null(x)) stop("GSEA returned NULL; inspect gene mapping and gene-set overlap.")
  if (nrow(x@result) == 0L) return(x)

  clusterProfiler::setReadable(
    x,
    OrgDb = org.Hs.eg.db,
    keyType = "ENTREZID"
  )
}

run_gse_go_SEA <- function(gene_list, ont) {
  x <- clusterProfiler::gseGO(
    geneList = gene_list,
    OrgDb = org.Hs.eg.db,
    keyType = "ENTREZID",
    ont = ont,
    minGSSize = minimum_pathway_size_SEA,
    maxGSSize = maximum_pathway_size_SEA,
    pvalueCutoff = 1,
    pAdjustMethod = "BH",
    eps = 0,
    seed = TRUE,
    by = "fgsea",
    verbose = FALSE
  )

  make_readable_gsea_SEA(x)
}

run_gse_kegg_SEA <- function(gene_list) {
  x <- clusterProfiler::gseKEGG(
    geneList = gene_list,
    organism = "hsa",
    keyType = "ncbi-geneid",
    minGSSize = minimum_pathway_size_SEA,
    maxGSSize = maximum_pathway_size_SEA,
    pvalueCutoff = 1,
    pAdjustMethod = "BH",
    eps = 0,
    seed = TRUE,
    by = "fgsea",
    verbose = FALSE
  )

  make_readable_gsea_SEA(x)
}

run_for_contrasts_SEA <- function(fun, ...) {
  out <- lapply(ranked_lists_SEA, fun, ...)
  names(out) <- names(ranked_lists_SEA)
  out
}

gsea_results_SEA <- list(
  GO_BP = run_for_contrasts_SEA(run_gse_go_SEA, ont = "BP"),
  GO_MF = run_for_contrasts_SEA(run_gse_go_SEA, ont = "MF"),
  GO_CC = run_for_contrasts_SEA(run_gse_go_SEA, ont = "CC"),
  KEGG = run_for_contrasts_SEA(run_gse_kegg_SEA)
)
```

# Export all contrasts



```{r full_tables}
for (database_name in names(gsea_results_SEA)) {
  for (contrast_name in names(gsea_results_SEA[[database_name]])) {
    x <- gsea_results_SEA[[database_name]][[contrast_name]]

    readr::write_csv(
      as.data.frame(x),
      file.path(
        table_folder_SEA,
        paste0(
          database_name,
          "_",
          contrast_name,
          "_all_GSEA_SEA.csv"
        )
      )
    )
  }
}
```

# Identify operational CT concepts



```{r conjunctional_concepts}
extract_contrast_result_SEA <- function(x, contrast_name) {
  result <- as.data.frame(x)

  if (nrow(result) == 0L) {
    result <- data.frame(ID = character(), Description = character(), setSize = integer(),
      NES = numeric(), pvalue = numeric(), p.adjust = numeric(), core_enrichment = character())
  }

  result %>%
    dplyr::select(
      ID,
      Description,
      setSize,
      NES,
      pvalue,
      p.adjust,
      core_enrichment
    ) %>%
    dplyr::rename_with(
      ~paste0(.x, "_", contrast_name),
      -ID
    )
}

find_conjunctional_concepts_SEA <- function(result_list) {
  r_only <- extract_contrast_result_SEA(result_list$R, "R")
  u_only <- extract_contrast_result_SEA(result_list$U, "U")
  combo <- extract_contrast_result_SEA(result_list$RU, "RU")

  dplyr::full_join(r_only, u_only, by = "ID") %>%
    dplyr::full_join(combo, by = "ID") %>%
    dplyr::filter(
      !is.na(p.adjust_RU),
      p.adjust_RU < adjusted_p_cutoff_SEA,
      is.na(p.adjust_R) | p.adjust_R >= adjusted_p_cutoff_SEA,
      is.na(p.adjust_U) | p.adjust_U >= adjusted_p_cutoff_SEA
    ) %>%
    dplyr::arrange(p.adjust_RU, dplyr::desc(abs(NES_RU)))
}

conjunctional_concepts_SEA <- lapply(
  gsea_results_SEA,
  find_conjunctional_concepts_SEA
)

concept_summary_SEA <- tibble::tibble(
  Database = names(conjunctional_concepts_SEA),
  n_conjunctional_concepts = vapply(
    conjunctional_concepts_SEA,
    nrow,
    integer(1)
  )
)

for (database_name in names(conjunctional_concepts_SEA)) {
  readr::write_csv(
    conjunctional_concepts_SEA[[database_name]],
    file.path(
      table_folder_SEA,
      paste0(
        database_name,
        "_operational_conjunctional_targeting_SEA.csv"
      )
    )
  )
}

readr::write_csv(
  concept_summary_SEA,
  file.path(table_folder_SEA, "conjunctional_targeting_summary_SEA.csv")
)

print(concept_summary_SEA)
```

# Prepare network inputs and save results



```{r save_biology}
# Select plotting inputs and calculate term similarities here, once.
plot_inputs_SEA <- lapply(names(gsea_results_SEA), function(database_name) {
  result_list <- gsea_results_SEA[[database_name]]
  selected <- lapply(result_list, function(x) {
    x@result <- x@result[!is.na(x@result$p.adjust) &
                          x@result$p.adjust < adjusted_p_cutoff_SEA, , drop = FALSE]
    x
  })
  ct <- result_list$RU
  ct@result <- ct@result[ct@result$ID %in% conjunctional_concepts_SEA[[database_name]]$ID, , drop = FALSE]
  selected$CT <- ct
  lapply(selected, function(x) {
    if (nrow(x@result) >= 2L) {
      x <- enrichplot::pairwise_termsim(x, method = "JC",
        showCategory = min(200L, nrow(x@result)))
    }
    x
  })
})
names(plot_inputs_SEA) <- names(gsea_results_SEA)
saveRDS(list(gsea_results = gsea_results_SEA, conjunctional_concepts = conjunctional_concepts_SEA,
            ranked_lists = ranked_lists_SEA), file.path(paths_SEA$data, "Biological_Concept_results_SEA.rds"))
saveRDS(list(plot_inputs = plot_inputs_SEA, show_category = show_category_SEA,
            minimum_term_similarity = minimum_term_similarity_SEA),
        file.path(paths_SEA$data, "Biological_Concept_plot_data_SEA.rds"))
writeLines(c(paste("Run date:", Sys.Date()),
  "GSEA: fgsea; GO: installed org.Hs.eg.db; KEGG: live hsa gene sets",
  "Gene-set size 10-500; BH-adjusted P < 0.10 within database and contrast",
  "Operational CT: RU significant; R and U nonsignificant or unavailable",
  "Term similarity: Jaccard; minimum edge similarity 0.20",
  "Seed: 20260908"), file.path(paths_SEA$root, "analysis_parameters_SEA.txt"))
save_session_SEA(paths_SEA$root)
```
