---
title: "5.1 Base and PTM CT proteins"
output:
  html_document:
    toc: true
    number_sections: true
date: "`r format(Sys.Date())`"
---

Run after 3.1 in the SEA series. Compare Base and PTM CT proteins using the same definition: `RU_sig & !Ru_sig & !Ul_sig`. Protein calls retain raw P < 0.05 and absolute median log2 fold change > 1 in at least one mapped peptide. Match exact protein accessions; do not merge isoforms or protein groups. The Base results are read from the existing `Limma/data/Limma_results.rds`.

```{r setup}
source("CoPPRA_config_SEA.R", local = TRUE)
source("CoPPRA_CT_comparison.R", local = TRUE)
library(ggplot2)
```

# Read both protein results

```{r read_results}
paths_SEA <- stage_paths_SEA("Comparisons/CT_Base_vs_SEA")
table_folder_SEA <- paths_SEA$tables
plot_folder_SEA <- paths_SEA$plots
original_dir <- path.expand(Sys.getenv(
  "COPPRA_BASE_OUTPUT", unset = file.path(dirname(project_dir_SEA), "Base", "Results")
))
original_file <- file.path(original_dir, "Limma", "data", "Limma_results.rds")
sea_file <- file.path(output_root_SEA, "Limma", "data", "Limma_results_SEA.rds")
original <- readRDS(original_file)
results_SEA <- readRDS(sea_file)

# Compare like-for-like thresholds and use the saved protein calls directly.
stopifnot(identical(original$pval_threshold, results_SEA$pval_threshold),
          identical(original$log2fc_threshold, results_SEA$log2fc_threshold))
protein <- as.data.table(original$protein_significance)
protein_SEA <- as.data.table(results_SEA$protein_significance)
stopifnot(!anyDuplicated(protein$ProteinID), !anyDuplicated(protein_SEA$ProteinID),
          !anyNA(protein[, .(ProteinID, Ru_sig, Ul_sig, RU_sig)]),
          !anyNA(protein_SEA[, .(ProteinID, Ru_sig, Ul_sig, RU_sig)]))
ct <- protein[RU_sig & !Ru_sig & !Ul_sig]
ct_SEA <- protein_SEA[RU_sig & !Ru_sig & !Ul_sig]
setorder(ct, ProteinID)
setorder(ct_SEA, ProteinID)
```

# CT proteins and intersections

```{r ct_tables}
shared_SEA <- intersect(ct$ProteinID, ct_SEA$ProteinID)
original_only <- setdiff(ct$ProteinID, ct_SEA$ProteinID)
sea_only <- setdiff(ct_SEA$ProteinID, ct$ProteinID)

membership_SEA <- merge(
  ct[, .(ProteinID, Base = TRUE, Base_groups = ProteinGroups)],
  ct_SEA[, .(ProteinID, SEA = TRUE, SEA_groups = ProteinGroups)],
  by = "ProteinID", all = TRUE
)
membership_SEA[is.na(Base), Base := FALSE]
membership_SEA[is.na(SEA), SEA := FALSE]
membership_SEA[, Intersection := fifelse(Base & SEA, "Shared",
  fifelse(Base, "Base only", "SEA only"))]
setorder(membership_SEA, ProteinID)

counts_SEA <- data.table(
  Set = c("Base CT", "SEA CT", "Shared CT", "Base only",
          "SEA only", "Union"),
  Proteins = c(nrow(ct), nrow(ct_SEA), length(shared_SEA), length(original_only),
               length(sea_only), nrow(membership_SEA))
)
stopifnot(nrow(ct) == length(original_only) + length(shared_SEA),
          nrow(ct_SEA) == length(sea_only) + length(shared_SEA),
          nrow(membership_SEA) == length(original_only) + length(shared_SEA) + length(sea_only))

fwrite(ct, file.path(table_folder_SEA, "CT_Base.csv"))
fwrite(ct_SEA, file.path(table_folder_SEA, "CT_SEA.csv"))
fwrite(membership_SEA[Intersection == "Shared"],
       file.path(table_folder_SEA, "CT_shared_SEA.csv"))
fwrite(ct[ProteinID %in% original_only],
       file.path(table_folder_SEA, "CT_Base_only_SEA.csv"))
fwrite(ct_SEA[ProteinID %in% sea_only],
       file.path(table_folder_SEA, "CT_SEA_only.csv"))
write_tables_SEA(list(CT_membership = membership_SEA, CT_counts = counts_SEA), table_folder_SEA)
knitr::kable(counts_SEA, caption = "CT protein counts and intersections")
```

# CT protein Venn diagram

The labels show exact counts. The two circles are schematic and their areas are not proportional to the set sizes.

```{r venn_plot, fig.width=8, fig.height=6}
p_venn_SEA <- ct_venn_plot(list(Base = ct$ProteinID, SEA = ct_SEA$ProteinID))

print(p_venn_SEA)
save_plot_SEA(p_venn_SEA, "CT_Venn", 8, 6)
```

# Protein lists

```{r protein_lists}
knitr::kable(ct, caption = "Base CT proteins")
knitr::kable(ct_SEA, caption = "SEA CT proteins")
knitr::kable(membership_SEA[Intersection == "Shared"], caption = "Shared CT proteins")
knitr::kable(ct[ProteinID %in% original_only], caption = "CT proteins found only in Base")
knitr::kable(ct_SEA[ProteinID %in% sea_only], caption = "CT proteins found only in SEA")
```

# Save comparison

```{r save_ct}
saveRDS(list(Base = ct, SEA = ct_SEA, membership = membership_SEA,
             counts = counts_SEA), file.path(paths_SEA$data, "CT_results_SEA.rds"))
fwrite(data.table(Dataset = c("Base", "SEA"),
                  File = c(original_file, sea_file),
                  MD5 = unname(tools::md5sum(c(original_file, sea_file)))),
       file.path(table_folder_SEA, "CT_input_files_SEA.csv"))
save_session_SEA(paths_SEA$root)
```
