---
title: "CT proteins: Base versus PTM v2"
output:
  html_document:
    toc: true
date: "`r format(Sys.Date())`"
---

Compare the current saved limma results for `evidence_new.txt` and `evidence_SEA_v2.txt`. CT proteins satisfy `RU_sig & !Ru_sig & !Ul_sig`. Match exact protein accessions without merging isoforms. This report reads the existing analysis; it does not rerun QC or limma.

```{r setup, message=FALSE, warning=FALSE}
source("CoPPRA_config_SEA.R", local = TRUE)
source("CoPPRA_CT_comparison.R", local = TRUE)
source("CoPPRA_validation_SEA.R", local = TRUE)
library(ggplot2)
library(jsonlite)
validation_theme_SEA()
comparison_paths <- stage_paths_SEA("Comparisons/CT_Base_vs_SEA")
plot_folder_SEA <- comparison_paths$plots
out_dir <- comparison_paths$tables
base_output <- path.expand(Sys.getenv("COPPRA_BASE_OUTPUT",
  unset = file.path(dirname(project_dir_SEA), "Base", "Results")))
new_file <- file.path(base_output, "Limma/data/Limma_results.rds")
sea_file <- file.path(output_root_SEA, "Limma/data/Limma_results_SEA.rds")
new <- readRDS(new_file)
sea <- readRDS(sea_file)
stopifnot(identical(new$pval_threshold, sea$pval_threshold),
          identical(new$log2fc_threshold, sea$log2fc_threshold))

get_ct <- function(result) {
  x <- as.data.table(result$protein_significance)
  stopifnot(!anyDuplicated(x$ProteinID),
            !anyNA(x[, .(ProteinID, Ru_sig, Ul_sig, RU_sig)]))
  sort(x[RU_sig & !Ru_sig & !Ul_sig, ProteinID])
}
ct_new <- get_ct(new)
ct_sea <- get_ct(sea)
common_ids <- intersect(ct_new, ct_sea)
only_new <- setdiff(ct_new, ct_sea)
only_sea <- setdiff(ct_sea, ct_new)
counts <- data.table(
  Set = c("Base CT", "SEA CT", "Common CT",
          "Base only", "SEA only", "Union"),
  Proteins = c(length(ct_new), length(ct_sea), length(common_ids),
              length(only_new), length(only_sea), length(union(ct_new, ct_sea))))
stopifnot(length(ct_new) == length(common_ids) + length(only_new),
          length(ct_sea) == length(common_ids) + length(only_sea))

clean_plot_text_SEA <- function(p) {
  p +
    labs(subtitle = NULL, caption = NULL, tag = NULL) +
    theme(
      plot.subtitle = element_blank(),
      plot.caption = element_blank(),
      plot.tag = element_blank()
    )
}
```

# CT protein counts

```{r counts}
knitr::kable(counts, caption = "CT protein counts and overlap")
fwrite(counts, file.path(out_dir, "CT_Base_vs_SEA_counts.csv"))
```

# Venn plot

Circle areas are schematic; labels give the exact counts. The PNG is saved at 600 dpi (4800 × 3600 pixels); a vector PDF is also saved.

```{r venn, fig.width=8, fig.height=6}
p <- venn_validation(list(Base = ct_new, SEA = ct_sea), "CT protein overlap")
p <- clean_plot_text_SEA(p)
print(p);save_plot_SEA(p, "CT_Base_vs_SEA_Venn", 3, 3, dpi = 600)
```

# Common CT proteins

Gene names retain the source peptide annotations. Peptide counts are distinct sequences in the saved limma table; RU-significant peptide counts use the existing `RU_sig` calls.

```{r common_proteins}
annotate_ct <- function(result, ids, tag) {
  pep <- as.data.table(result$peptide_significance)
  mapping <- pep[!is.na(Proteins) & nzchar(Proteins),
    .(ProteinID = trimws(unlist(strsplit(Proteins, ";", fixed = TRUE)))),
    by = .(Sequence, Gene_Names, RU_sig)]
  mapping <- mapping[ProteinID %in% ids]
  annotation <- mapping[, .(
    Gene_names = paste(sort(unique(na.omit(Gene_Names[nzchar(Gene_Names)]))), collapse = ";"),
    Peptides = uniqueN(Sequence), RU_significant_peptides = uniqueN(Sequence[RU_sig])), by = ProteinID]
  setnames(annotation, c("Gene_names", "Peptides", "RU_significant_peptides"),
    paste0(c("Gene_names_", "Peptides_", "RU_significant_peptides_"), tag))
  annotation
}
common <- merge(data.table(ProteinID = common_ids), annotate_ct(new, common_ids, "new"),
                by = "ProteinID", all.x = TRUE)
common <- merge(common, annotate_ct(sea, common_ids, "SEA"), by = "ProteinID", all.x = TRUE)
setcolorder(common, c("ProteinID", "Gene_names_new", "Gene_names_SEA", "Peptides_new",
  "Peptides_SEA", "RU_significant_peptides_new", "RU_significant_peptides_SEA"))
setorder(common, ProteinID)
stopifnot(nrow(common) == length(common_ids), !anyDuplicated(common$ProteinID))
knitr::kable(common, caption = "Common CT protein accessions")
```

# Excel report

The workbook contains a counts sheet and a filterable common-protein table. The export uses the installed Codex spreadsheet runtime; override `COPPRA_ARTIFACT_RUNTIME` if that runtime moves.

```{r excel_export, results='hide', message=FALSE, warning=FALSE}
runtime <- path.expand(Sys.getenv("COPPRA_ARTIFACT_RUNTIME",
  unset = "~/.cache/codex-runtimes/codex-primary-runtime/dependencies/node"))
node <- file.path(runtime, "bin/node")
stopifnot(file.exists(node), dir.exists(file.path(runtime, "node_modules/@oai/artifact-tool")))
tmp <- tempfile("ct_excel_")
dir.create(tmp)
stopifnot(file.symlink(file.path(runtime, "node_modules"), file.path(tmp, "node_modules")))
payload <- list(counts = counts, common = common,
  sources = data.table(Dataset = c("Base/evidence_new.txt", "SEA/evidence_SEA_v2.txt"),
    Result = c(new_file, sea_file),
    MD5 = unname(tools::md5sum(c(new_file, sea_file)))),
  rule = "RU_sig & !Ru_sig & !Ul_sig; exact protein accession matching",
  columns = names(common))
write_json(payload, file.path(tmp, "data.json"), dataframe = "rows", auto_unbox = TRUE, na = "null")
js <- r"---(
import fs from 'node:fs/promises';
import {Workbook, SpreadsheetFile} from '@oai/artifact-tool';
const d = JSON.parse(await fs.readFile(process.argv[2], 'utf8'));
const wb = Workbook.create();
const s = wb.worksheets.add('CT counts');
const c = wb.worksheets.add('Common CT proteins');
for (const sh of [s,c]) sh.showGridLines = false;
s.getRange('A2').values = [['CT protein comparison']];
s.getRange('A4:B10').values = [['Set','Proteins'], ...d.counts.map(x=>[x.Set,x.Proteins])];
s.getRange('A13').values = [['Definition']];
s.getRange('A14').values = [[d.rule]];
s.getRange('A17:C19').values = [['Input dataset','Saved result','MD5'], ...d.sources.map(x=>[x.Dataset,x.Result,x.MD5])];
s.getRange('A2:C19').format.font = {name:'Arial',size:11};
s.getRange('A2').format.font = {name:'Arial',size:16,bold:true};
s.getRange('A4:B4').format = {fill:'#234E70',font:{name:'Arial',bold:true,color:'#FFFFFF'}};
s.getRange('A17:C17').format = {fill:'#234E70',font:{name:'Arial',bold:true,color:'#FFFFFF'}};
s.getRange('A2:A19').format.columnWidth = 33;
s.getRange('B2:B19').format.columnWidth = 53;
s.getRange('C2:C19').format.columnWidth = 39;
s.getRange('A4:C19').format.rowHeight = 23;
s.getRange('B5:B10').setNumberFormat('#,##0');
const headers = ['ProteinID','Gene names: Base','Gene names: SEA',
 'Peptides: Base','Peptides: SEA','RU-significant peptides: Base','RU-significant peptides: SEA'];
const rows = d.common.map(x=>d.columns.map(k=>x[k]??null));
const n = rows.length+1;
c.getRange(`A1:G${n}`).values = [headers,...rows];
c.getRange(`A1:G${n}`).format.font = {name:'Arial',size:11};
c.getRange(`A1:G${n}`).format.rowHeight = 22;
c.getRange(`A1:A${n}`).format.columnWidth = 18;
c.getRange(`B1:C${n}`).format.columnWidth = 29;
c.getRange(`D1:E${n}`).format.columnWidth = 24;
c.getRange(`F1:G${n}`).format.columnWidth = 27;
c.getRange('A1:G1').format = {fill:'#234E70',font:{name:'Arial',bold:true,color:'#FFFFFF'},wrapText:true,rowHeight:43};
if(n>1) { c.getRange(`D2:G${n}`).setNumberFormat('#,##0'); c.tables.add(`A1:G${n}`,true,'CommonCTProteins'); }
c.freezePanes.freezeRows(1);
c.freezePanes.freezeColumns(1);
wb.recalculate();
console.log((await wb.inspect({kind:'table',range:"'CT counts'!A4:B10",include:'values',tableMaxRows:7,tableMaxCols:2})).ndjson);
for (const [sheetName,range,name] of [['CT counts','A2:C19','counts'],['Common CT proteins','A1:G12','common']]) {
 const preview = await wb.render({sheetName,range,scale:1,format:'png'});
 await fs.writeFile(`${process.argv[4]}/${name}.png`,new Uint8Array(await preview.arrayBuffer()));
}
const file = await SpreadsheetFile.exportXlsx(wb);
await file.save(process.argv[3]);
await fs.rename(`${process.argv[3]}.inspect.ndjson`, `${process.argv[4]}/workbook.inspect.ndjson`).catch(()=>{});
)---"
writeLines(js, file.path(tmp, "workbook.mjs"))
preview_dir <- Sys.getenv("COPPRA_CT_PREVIEW_DIR", unset = tmp)
dir.create(preview_dir, recursive = TRUE, showWarnings = FALSE)
xlsx_file <- file.path(out_dir, "CT_common_Base_vs_SEA.xlsx")
status <- system2(node, shQuote(c(file.path(tmp, "workbook.mjs"), file.path(tmp, "data.json"),
  xlsx_file, preview_dir)))
stopifnot(status == 0L, file.exists(xlsx_file))
```

Saved workbook: `CT_common_Base_vs_SEA.xlsx` in `Results_v2/Comparisons/CT_Base_vs_SEA/tables`.
