---
title: "3.1 CoPPRA limma analysis"
output:
  html_document:
    toc: true
    number_sections: true
date: "`r format(Sys.Date())`"
---

Run after 1.1 (or in the numbered series). Carriers are excluded. The saved normalized matrix is used directly. As in v3, missing values are replaced with each peptide median only in the limma input; median effect sizes use observed biological values. Calls retain raw P < 0.05 and absolute median log2 fold change > 1. BH-adjusted P values are also exported. Protein calls mean at least one mapped peptide passes; RU-only calls require `RU_sig & !Ru_sig & !Ul_sig`.

```{r setup}
source("CoPPRA_config.R", local = TRUE)
library(dplyr)
library(tidyr)
library(matrixStats)
library(limma)
```


# Read shared quantification



```{r read_quantification}
paths <- stage_paths("Limma")
table_folder <- paths$tables
pval_threshold <- 0.05
log2fc_threshold <- log2(2)
q <- readRDS(file.path(output_root, "QC", "data", "quantification.rds"))
analysis_meta <- copy(q$sample_map[Condition != "Carrier"])
model_labels <- c(Ctrl = "Control", R = "Ru", U = "Ul", RU = "RU")
analysis_meta[, Condition := factor(model_labels[as.character(Condition)],
                                    levels = c("Control", "Ru", "Ul", "RU"))]
setorder(analysis_meta, Condition, Bio_rep)
analysis_samples <- analysis_meta$Experiment
peptide_annotation <- q$peptide_annotation
peptide_matrix <- q$biological_log2[, analysis_samples, drop = FALSE]
fwrite(analysis_meta, file.path(table_folder, "differential_sample_metadata.csv"))
```

# Fit limma contrasts



```{r limma_analysis}
row_medians <- rowMedians(peptide_matrix, na.rm = TRUE)
finite_rows <- is.finite(row_medians)

peptide_matrix <- peptide_matrix[finite_rows, , drop = FALSE]
row_medians <- row_medians[finite_rows]

limma_matrix <- peptide_matrix
missing_index <- which(!is.finite(limma_matrix), arr.ind = TRUE)
limma_matrix[missing_index] <- row_medians[missing_index[, 1]]

condition <- analysis_meta$Condition
design <- model.matrix(~ 0 + condition)
colnames(design) <- levels(condition)
rownames(design) <- analysis_samples

contrast_matrix <- makeContrasts(
  Ru_vs_Control = Ru - Control,
  Ul_vs_Control = Ul - Control,
  RU_vs_Control = RU - Control,
  levels = design
)

fit <- lmFit(limma_matrix, design)
fit <- contrasts.fit(fit, contrast_matrix)
fit <- eBayes(fit)

peptide_results <- data.table(
  Sequence = rownames(limma_matrix)
)

contrast_results <- lapply(colnames(contrast_matrix), function(contrast_name) {
  result <- as.data.table(
    topTable(
      fit,
      coef = contrast_name,
      number = Inf,
      sort.by = "none",
      adjust.method = "BH"
    ),
    keep.rownames = "Sequence"
  )

  result <- result[, .(Sequence, logFC, P.Value, adj.P.Val)]
  setnames(
    result,
    c("logFC", "P.Value", "adj.P.Val"),
    paste0(
      contrast_name,
      c("_limma_logFC", "_pvalue", "_adj.P.Val")
    )
  )
  result
})

peptide_results <- Reduce(
  function(x, y) merge(x, y, by = "Sequence"),
  c(list(peptide_results), contrast_results)
)

fwrite(
  peptide_results,
  file.path(table_folder, "peptide_limma_results.csv")
)

cat("Total peptides tested:", nrow(peptide_results), "\n")
```

# Calculate median effects and peptide calls



```{r peptide_significance}
condition_samples <- split(analysis_meta$Experiment, analysis_meta$Condition)
condition_medians <- lapply(condition_samples, function(samples) {
  matrixStats::rowMedians(peptide_matrix[, samples, drop = FALSE], na.rm = TRUE)
})
peptide_effects <- data.table(Sequence = rownames(peptide_matrix))
for (contrast in c("Ru", "Ul", "RU")) {
  peptide_effects[, (paste0(contrast, "_log2FC")) :=
    condition_medians[[contrast]] - condition_medians$Control]
}
peptide_significance <- merge(peptide_results, peptide_effects, by = "Sequence")
peptide_significance <- merge(peptide_significance, peptide_annotation, by = "Sequence", all.x = TRUE)
for (contrast in c("Ru", "Ul", "RU")) {
  p_value <- peptide_significance[[paste0(contrast, "_vs_Control_pvalue")]]
  effect <- peptide_significance[[paste0(contrast, "_log2FC")]]
  peptide_significance[, (paste0(contrast, "_sig")) :=
    is.finite(p_value) & is.finite(effect) &
    p_value < pval_threshold & abs(effect) > log2fc_threshold]
}
significant_peptides <- peptide_significance[Ru_sig | Ul_sig | RU_sig]
write_tables(list(all_peptide_differential_results = peptide_significance,
                 significant_peptides = significant_peptides,
                 peptide_biological_matrix = matrix_table(peptide_matrix),
                 peptide_limma_matrix_used = matrix_table(limma_matrix)), table_folder)
```

# Collapse peptide calls to protein accessions



```{r protein_collapse}
protein_significance <- peptide_significance %>%
  dplyr::as_tibble() %>%
  dplyr::filter(!is.na(Proteins), Proteins != "") %>%
  dplyr::mutate(ProteinGroup = Proteins) %>%
  tidyr::separate_rows(Proteins, sep = ";") %>%
  dplyr::mutate(ProteinID = trimws(Proteins)) %>%
  dplyr::filter(ProteinID != "") %>%
  dplyr::distinct(Sequence, ProteinID, .keep_all = TRUE) %>%
  dplyr::group_by(ProteinID) %>%
  dplyr::summarise(
    Ru_sig = any(Ru_sig, na.rm = TRUE),
    Ul_sig = any(Ul_sig, na.rm = TRUE),
    RU_sig = any(RU_sig, na.rm = TRUE),
    ProteinGroups = paste(unique(ProteinGroup), collapse = " | "),
    .groups = "drop"
  ) %>%
  dplyr::filter(Ru_sig | Ul_sig | RU_sig)

conjunctional_proteins <- protein_significance %>%
  dplyr::filter(RU_sig & !Ru_sig & !Ul_sig)

fwrite(
  as.data.table(protein_significance),
  file.path(table_folder, "significant_proteins.csv")
)
fwrite(
  as.data.table(conjunctional_proteins),
  file.path(table_folder, "conjunctional_targeting_proteins.csv")
)

protein_count_summary <- data.table(
  Set = c("R", "U", "RU", "RU only / conjunctional"),
  Significant_proteins = c(
    sum(protein_significance$Ru_sig),
    sum(protein_significance$Ul_sig),
    sum(protein_significance$RU_sig),
    nrow(conjunctional_proteins)
  )
)

knitr::kable(
  protein_count_summary,
  caption = "Significant proteins by set"
)
```

# Calculate intersection tables



```{r intersection_tables}
make_intersection_table <- function(data, set_columns) {
  data %>%
    dplyr::select(ProteinID, dplyr::all_of(set_columns)) %>%
    dplyr::filter(
      dplyr::if_any(dplyr::all_of(set_columns), identity)
    ) %>%
    tidyr::pivot_longer(
      cols = -ProteinID,
      names_to = "Comparison",
      values_to = "Significant"
    ) %>%
    dplyr::filter(Significant) %>%
    dplyr::mutate(Comparison = sub("_sig", "", Comparison)) %>%
    dplyr::group_by(ProteinID) %>%
    dplyr::summarise(
      Sets = paste(sort(Comparison), collapse = "_"),
      .groups = "drop"
    ) %>%
    dplyr::group_by(Sets) %>%
    dplyr::summarise(
      n_proteins = dplyr::n(),
      Proteins = paste(ProteinID, collapse = ";"),
      .groups = "drop"
    )
}

protein_intersections <- make_intersection_table(
  protein_significance,
  c("Ru_sig", "RU_sig", "Ul_sig")
)

fwrite(
  as.data.table(protein_intersections),
  file.path(table_folder, "significant_intersections.csv")
)

knitr::kable(
  protein_intersections,
  caption = "Protein intersections represented in the UpSet plot"
)
```

# Save results



```{r save_limma}
analysis_summary <- data.table(
  Metric = c("Evidence rows", "Peptides tested by limma", "Significant peptides",
             "Significant proteins", "RU-only conjunctional proteins"),
  Value = c(q$n_evidence, nrow(peptide_results), nrow(significant_peptides),
            nrow(protein_significance), nrow(conjunctional_proteins))
)
write_tables(list(differential_analysis_summary = analysis_summary,
                 significant_protein_counts = protein_count_summary), table_folder)
saveRDS(list(peptide_significance = peptide_significance,
  protein_significance = protein_significance,
  pval_threshold = pval_threshold, log2fc_threshold = log2fc_threshold),
  file.path(paths$data, "Limma_results.rds"))
saveRDS(fit, file.path(paths$data, "limma_fit.rds"))
save_session(paths$root)
knitr::kable(analysis_summary)
```
