---
title: "7.1 CoPPRA topGO biological concept analysis"
output:
  html_document:
    toc: true
    number_sections: true
date: "`r format(Sys.Date())`"
---

Run independently after 3.1 using `Limma/data/Limma_results.rds` and the adjacent `CoPPRA_config.R`. Rebuild R, U and RU ranked lists with the same scoring and mapping as 4.1, obtain GO annotations from the installed databases, and run KEGG GSEA here. No 4.1 result or previous R session is required. KEGG requires internet access. Test GO BP/MF/CC with classic topGO KS in both rank directions; combine the two one-sided P values as min(1, 2 * min(P_up, P_down)), then apply BH within each ontology and contrast. Operational CT concepts pass adjusted P < 0.10 in RU and are nonsignificant or unavailable in R and U. This is an operational comparison, not an interaction test. The GO method comparison was exploratory. Packages must be installed before running the series.

```{r setup}
source("CoPPRA_config.R", local = TRUE)
library(dplyr)
library(readr)
library(AnnotationDbi)
library(clusterProfiler)
library(org.Hs.eg.db)
library(GO.db)
library(topGO)
library(enrichplot)
```

# Read ranked inputs and set parameters

```{r read_results}
paths <- stage_paths("TopGO")
table_folder <- paths$tables
minimum_pathway_size <- 15L
maximum_pathway_size <- 500L
minimum_reported_size <- 6L
adjusted_p_cutoff <- 0.10
minimum_term_similarity <- 0.20
show_category <- 12L
graph_node_size <- 5L
set.seed(2026)

input_file <- file.path(output_root, "Limma", "data", "Limma_results.rds")
if (!file.exists(input_file)) stop("Missing saved Limma input: ", input_file,
  ". Run 3.1 once before starting 7.1 independently.")
input_hash <- unname(tools::md5sum(input_file))
peptide_results <- data.table::as.data.table(
  data.table::copy(readRDS(input_file)$peptide_significance))
required_columns <- c("Sequence", "Proteins",
  "Ru_vs_Control_limma_logFC", "Ru_vs_Control_pvalue",
  "Ul_vs_Control_limma_logFC", "Ul_vs_Control_pvalue",
  "RU_vs_Control_limma_logFC", "RU_vs_Control_pvalue")
missing_columns <- setdiff(required_columns, names(peptide_results))
if (length(missing_columns)) stop("Limma input is missing columns: ",
  paste(missing_columns, collapse = ", "))
```

# Calculate protein scores and map UniProt to Entrez IDs

The scoring, accession aggregation, mapping and deterministic tie-breaking are
identical to 4.1; enrichment does not read or execute that stage.

```{r protein_scores}
signed_z <- function(log_fc, p_value) {
  # Keep unavailable statistics unavailable; only clamp finite P values at zero.
  p_value <- pmin(pmax(p_value, .Machine$double.xmin), 1)
  sign(log_fc) * stats::qnorm(p_value / 2, lower.tail = FALSE)
}

peptide_results[, `:=`(
  Ru_score = signed_z(
    Ru_vs_Control_limma_logFC,
    Ru_vs_Control_pvalue
  ),
  Ul_score = signed_z(
    Ul_vs_Control_limma_logFC,
    Ul_vs_Control_pvalue
  ),
  RU_score = signed_z(
    RU_vs_Control_limma_logFC,
    RU_vs_Control_pvalue
  )
)]

protein_score_long <- peptide_results[
  !is.na(Proteins) & Proteins != "",
  .(
    ProteinID = trimws(unlist(strsplit(
      Proteins,
      ";",
      fixed = TRUE
    )))
  ),
  by = .(
    Sequence,
    Ru_score,
    Ul_score,
    RU_score
  )
]

protein_score_long <- unique(protein_score_long[ProteinID != ""], by = c("Sequence", "ProteinID"))

protein_scores <- protein_score_long[, .(
  n_peptides = uniqueN(Sequence),
  R_score = median(Ru_score, na.rm = TRUE),
  U_score = median(Ul_score, na.rm = TRUE),
  RU_score = median(RU_score, na.rm = TRUE)
), by = ProteinID]

protein_scores[, UniProt := sub("-\\d+$", "", ProteinID)]

cat("Protein accessions with ranking scores:", nrow(protein_scores), "\n")
```

```{r protein_mapping}
uniprot_ids <- unique(protein_scores$UniProt)

id_map <- clusterProfiler::bitr(
  uniprot_ids,
  fromType = "UNIPROT",
  toType = c("ENTREZID", "SYMBOL"),
  OrgDb = org.Hs.eg.db
) %>%
  dplyr::distinct(UNIPROT, ENTREZID, SYMBOL)

protein_scores_mapped <- protein_scores %>%
  dplyr::inner_join(
    id_map,
    by = c("UniProt" = "UNIPROT")
  ) %>%
  dplyr::group_by(ENTREZID) %>%
  dplyr::summarise(
    SYMBOL = paste(sort(unique(SYMBOL)), collapse = ";"),
    UniProt = paste(sort(unique(UniProt)), collapse = ";"),
    n_accessions = dplyr::n_distinct(ProteinID),
    n_peptides = sum(n_peptides),
    R_score = median(R_score, na.rm = TRUE),
    U_score = median(U_score, na.rm = TRUE),
    RU_score = median(RU_score, na.rm = TRUE),
    .groups = "drop"
  )

readr::write_csv(
  id_map,
  file.path(table_folder, "uniprot_to_entrez_mapping.csv")
)

readr::write_csv(
  protein_scores_mapped,
  file.path(table_folder, "protein_rank_scores.csv")
)

mapping_summary <- tibble::tibble(
  Metric = c(
    "Protein accessions before mapping",
    "Mapped Entrez IDs used for ranking"
  ),
  Value = c(
    nrow(protein_scores),
    nrow(protein_scores_mapped)
  )
)

readr::write_csv(
  mapping_summary,
  file.path(table_folder, "ranking_mapping_summary.csv")
)

print(mapping_summary)
```

# Build ranked lists

```{r ranked_lists}
make_ranked_list <- function(data, score_column) {
  score <- data[[score_column]]
  names(score) <- as.character(data$ENTREZID)

  score <- score[is.finite(score) & !is.na(names(score))]

  tie_break <- rank(
    names(score),
    ties.method = "first"
  ) / length(score)

  score <- score + tie_break * 1e-12
  sort(score, decreasing = TRUE)
}

ranked_lists <- list(
  R = make_ranked_list(protein_scores_mapped, "R_score"),
  U = make_ranked_list(protein_scores_mapped, "U_score"),
  RU = make_ranked_list(protein_scores_mapped, "RU_score")
)

stopifnot(length(ranked_lists) == 3L,
  all(vapply(ranked_lists, function(x) is.numeric(x) && length(x) > 0L &&
    all(is.finite(x)) && !is.null(names(x)) && !anyNA(names(x)) &&
    all(nzchar(names(x))) && !anyDuplicated(names(x)), logical(1))))

rank_summary <- data.frame(Contrast = names(ranked_lists),
  n_ranked_genes = lengths(ranked_lists),
  minimum_score = vapply(ranked_lists, min, numeric(1)),
  maximum_score = vapply(ranked_lists, max, numeric(1)))
write_csv(rank_summary, file.path(table_folder, "ranked_list_summary.csv"))
for (contrast in names(ranked_lists)) {
  score <- ranked_lists[[contrast]]
  write_csv(data.frame(ENTREZID = names(score), Score = unname(score)),
    file.path(table_folder, paste0(contrast, "_ranked_background.csv")))
}
print(rank_summary)
```

# Prepare GO annotations directly from the installed databases

```{r go_annotations}
measured_genes <- unique(unlist(lapply(ranked_lists, names), use.names = FALSE))
go_annotation <- AnnotationDbi::select(org.Hs.eg.db,
  keys = measured_genes, keytype = "ENTREZID", columns = c("GOALL", "ONTOLOGYALL"))
# GOALL includes ancestor annotations, matching the gene sets used by gseGO.
go_annotation <- unique(go_annotation[!is.na(go_annotation$GOALL) &
  !is.na(go_annotation$ONTOLOGYALL), ])
go_ontology <- AnnotationDbi::Ontology(GO.db::GOTERM)
go_annotation <- go_annotation[go_annotation$GOALL %in% names(go_ontology), ]
go_gene_sets <- setNames(lapply(c("BP", "MF", "CC"), function(ontology) {
  annotation <- go_annotation[go_ontology[go_annotation$GOALL] == ontology, ]
  lapply(split(annotation$ENTREZID, annotation$GOALL), unique)
}), c("GO_BP", "GO_MF", "GO_CC"))
```

# Prepare the GO graph with the measured background

```{r go_graph}
make_go_data <- function(gene_list, gene_sets, ontology) {
  background <- names(gene_list)
  gene_sets <- lapply(gene_sets, intersect, y = background)
  gene_to_go <- split(rep(names(gene_sets), lengths(gene_sets)),
                      unlist(gene_sets, use.names = FALSE))
  root <- c(BP = "GO:0008150", MF = "GO:0003674", CC = "GO:0005575")[[ontology]]
  # Root membership keeps unannotated measured genes in the KS null universe.
  # It does not add them to any tested term; the root exceeds the size limit.
  gene_to_go <- setNames(lapply(background, function(gene) {
    unique(c(gene_to_go[[gene]], root))
  }), background)
  x <- new("topGOdata", ontology = ontology, allGenes = gene_list,
    geneSel = function(x) rep(FALSE, length(x)), nodeSize = graph_node_size,
    annot = annFUN.gene2GO, gene2GO = gene_to_go)
  stopifnot(setequal(genes(x), background))
  members <- genesInTerm(x)
  eligible <- names(members)[lengths(members) >= minimum_pathway_size &
                              lengths(members) <= maximum_pathway_size]
  # Validate that graph propagation has not changed any tested membership.
  stopifnot(all(vapply(eligible, function(id) {
    setequal(members[[id]], gene_sets[[id]])
  }, logical(1))))
  list(data = x, members = members[eligible])
}
```

# Run classic topGO KS for each contrast

```{r topgo_tests}
run_topgo <- function(gene_list, gene_sets, ontology) {
  prepared <- make_go_data(gene_list, gene_sets, ontology)
  x <- prepared$data
  members <- prepared$members
  up <- runTest(x, algorithm = "classic", statistic = "ks", scoreOrder = "decreasing")
  down <- runTest(x, algorithm = "classic", statistic = "ks", scoreOrder = "increasing")
  ids <- names(members)
  p_up <- score(up)[ids]
  p_down <- score(down)[ids]
  p_two <- pmin(1, 2 * pmin(p_up, p_down))
  direction <- ifelse(p_up <= p_down, "up", "down")
  ranks <- rank(-gene_list)
  # Signed KS distance is an effect descriptor, not a GSEA NES.
  signed_ks <- vapply(seq_along(ids), function(i) {
    inside <- ranks[members[[i]]]
    outside <- ranks[setdiff(names(ranks), members[[i]])]
    alternative <- if (direction[i] == "up") "greater" else "less"
    distance <- unname(stats::ks.test(inside, outside,
                                      alternative = alternative)$statistic)
    if (direction[i] == "up") distance else -distance
  }, numeric(1))
  annotation <- AnnotationDbi::select(GO.db, keys = ids,
                                      keytype = "GOID", columns = "TERM")
  result <- data.frame(ID = ids, Description = annotation$TERM[match(ids, annotation$GOID)],
    setSize = lengths(members), KSstatistic = signed_ks, Direction = direction,
    p_up = unname(p_up), p_down = unname(p_down), pvalue = unname(p_two),
    p.adjust = p.adjust(p_two, method = "BH"),
    geneID = vapply(members, paste, collapse = "/", character(1)),
    Count = lengths(members), stringsAsFactors = FALSE)
  result <- result[order(result$p.adjust, -abs(result$KSstatistic), result$ID), ]
  # An enrichResult is only a plotting container here; its P values are from topGO.
  # GeneRatio describes measured term membership, not a selected-hit ORA ratio.
  result$GeneRatio <- paste0(result$Count, "/", length(gene_list))
  result$qvalue <- NA_real_
  plot_result <- new("enrichResult", result = result, pvalueCutoff = 1,
    pAdjustMethod = "BH", qvalueCutoff = 1, organism = "Homo sapiens",
    ontology = ontology, gene = names(gene_list), keytype = "ENTREZID",
    universe = names(gene_list), geneSets = members, readable = FALSE)
  plot_result <- clusterProfiler::setReadable(plot_result,
                              OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
  list(result = plot_result, tests = list(up = up, down = down), table = result)
}

enrichment_results <- list()
topgo_tests <- list()
for (ontology in c("BP", "MF", "CC")) {
  database <- paste0("GO_", ontology)
  fits <- lapply(names(ranked_lists), function(contrast) {
    run_topgo(ranked_lists[[contrast]],
      go_gene_sets[[database]], ontology)
  })
  names(fits) <- names(ranked_lists)
  enrichment_results[[database]] <- lapply(fits, `[[`, "result")
  topgo_tests[[database]] <- lapply(fits, `[[`, "tests")
  for (contrast in names(fits)) {
    write_csv(fits[[contrast]]$table,
      file.path(table_folder, paste0(database, "_", contrast, "_all_topGO.csv")))
  }
}
```

# Run KEGG GSEA independently

```{r kegg_gsea}
run_gse_kegg <- function(gene_list) {
  x <- clusterProfiler::gseKEGG(
    geneList = gene_list, organism = "hsa", keyType = "ncbi-geneid",
    minGSSize = minimum_pathway_size, maxGSSize = maximum_pathway_size,
    pvalueCutoff = 1, pAdjustMethod = "BH", eps = 0,
    seed = TRUE, by = "fgsea", verbose = FALSE)
  if (is.null(x)) stop("KEGG GSEA returned NULL; inspect gene mapping and gene-set overlap.")
  if (nrow(x@result) == 0L) return(x)
  clusterProfiler::setReadable(x, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
}
enrichment_results$KEGG <- lapply(ranked_lists, run_gse_kegg)
for (contrast in names(ranked_lists)) {
  write_csv(as.data.frame(enrichment_results$KEGG[[contrast]]),
    file.path(table_folder, paste0("KEGG_", contrast, "_all_GSEA.csv")))
}
```

# Identify operational CT concepts

```{r conjunctional_concepts}
extract_contrast_result <- function(x, contrast) {
  result <- as.data.frame(x)
  # Keep method-specific effect columns: KSstatistic for GO and NES for KEGG.
  result %>%
    dplyr::select(dplyr::any_of(c("ID", "Description", "setSize", "KSstatistic",
      "Direction", "NES", "pvalue", "p.adjust", "geneID", "core_enrichment"))) %>%
    dplyr::rename_with(~paste0(.x, "_", contrast), -ID)
}

find_conjunctional_concepts <- function(result_list) {
  r_only <- extract_contrast_result(result_list$R, "R")
  u_only <- extract_contrast_result(result_list$U, "U")
  combo <- extract_contrast_result(result_list$RU, "RU")
  dplyr::full_join(r_only, u_only, by = "ID") %>%
    dplyr::full_join(combo, by = "ID") %>%
    dplyr::filter(!is.na(p.adjust_RU), p.adjust_RU < adjusted_p_cutoff,
      is.na(p.adjust_R) | p.adjust_R >= adjusted_p_cutoff,
      is.na(p.adjust_U) | p.adjust_U >= adjusted_p_cutoff,
      setSize_RU >= minimum_reported_size) %>%
    dplyr::arrange(p.adjust_RU, ID)
}

conjunctional_concepts <- lapply(enrichment_results, find_conjunctional_concepts)
concept_summary <- data.frame(Database = names(conjunctional_concepts),
  Method = c(rep("topGO classic KS", 3), "KEGG GSEA"),
  n_conjunctional_concepts = vapply(conjunctional_concepts, nrow, integer(1)))
for (database in names(conjunctional_concepts)) {
  write_csv(conjunctional_concepts[[database]],
    file.path(table_folder, paste0(database, "_operational_conjunctional_targeting.csv")))
}
write_csv(concept_summary, file.path(table_folder, "conjunctional_targeting_summary.csv"))
print(concept_summary)
```

# Prepare network inputs and save results

```{r save_biology}
plot_inputs <- lapply(names(enrichment_results), function(database) {
  result_list <- enrichment_results[[database]]
  selected <- lapply(result_list, function(x) {
    x@result <- x@result[!is.na(x@result$p.adjust) &
      x@result$p.adjust < adjusted_p_cutoff &
      x@result$setSize >= minimum_reported_size, , drop = FALSE]
    x
  })
  ct <- result_list$RU
  ct@result <- ct@result[ct@result$ID %in% conjunctional_concepts[[database]]$ID, , drop = FALSE]
  selected$CT <- ct
  lapply(selected, function(x) {
    if (nrow(x@result) >= 2L) {
      x <- enrichplot::pairwise_termsim(x, method = "JC",
                                       showCategory = min(200L, nrow(x@result)))
    }
    x
  })
})
names(plot_inputs) <- names(enrichment_results)
parameters <- list(minimum_pathway_size = minimum_pathway_size,
  maximum_pathway_size = maximum_pathway_size, minimum_reported_size = minimum_reported_size,
  adjusted_p_cutoff = adjusted_p_cutoff, graph_node_size = graph_node_size,
  GO_method = "topGO classic KS; two rank directions; 2*min(P); BH",
  KEGG_method = "KEGG GSEA run independently with fgsea; BH",
  GO_annotation = "org.Hs.eg.db GOALL; ontology from GO.db",
  org_Hs_eg_db_version = as.character(packageVersion("org.Hs.eg.db")),
  GO_db_version = as.character(packageVersion("GO.db")),
  input_file = input_file, input_md5 = input_hash)
saveRDS(list(enrichment_results = enrichment_results, topgo_tests = topgo_tests,
  conjunctional_concepts = conjunctional_concepts, ranked_lists = ranked_lists,
  parameters = parameters), file.path(paths$data, "TopGO_results.rds"))
saveRDS(list(plot_inputs = plot_inputs, show_category = show_category,
  minimum_term_similarity = minimum_term_similarity),
  file.path(paths$data, "TopGO_plot_data.rds"))
saveRDS(list(ranked_lists = ranked_lists, GO_gene_sets = go_gene_sets,
  KEGG_gene_sets = lapply(enrichment_results$KEGG, function(x) x@geneSets),
  input_file = input_file, input_md5 = input_hash),
  file.path(paths$data, "TopGO_input_data.rds"))
writeLines(c(paste("Run date:", Sys.Date()),
  "GO: topGO classic KS; both rank directions; min(1, 2*min(P_up, P_down))",
  "BH adjustment within each ontology and contrast; all size-eligible terms included",
  sprintf("Measured term size: %d-%d; reported size >= %d; adjusted P < %.2f",
    minimum_pathway_size, maximum_pathway_size, minimum_reported_size, adjusted_p_cutoff),
  "GO graph root membership retains every measured gene in the null universe",
  "GO plot size/network genes: all measured term members; effect: signed KS distance",
  "Rankings rebuilt from saved Limma results using the 4.1 scoring and mapping method",
  "GO annotation: installed org.Hs.eg.db GOALL and GO.db",
  "KEGG: independently calculated GSEA; plot effect: NES; network genes: leading edge",
  "Operational CT: RU significant; R and U nonsignificant or unavailable",
  "Method comparison is exploratory; operational CT is not an interaction test",
  paste("Input:", input_file), paste("Input MD5:", input_hash),
  paste("topGO version:", packageVersion("topGO")),
  "Term similarity: Jaccard; minimum edge similarity 0.20"),
  file.path(paths$root, "analysis_parameters.txt"))
save_session(paths$root)
```
